Signatures of Foreground Sky in Power Spectra of Redshifted Neutral Hydrogen from the Epoch of Reionization

### 2014-09-04 foreground_delay_spectrum_modeling.tex
interferometer replaced with baseline to match vernacular usage

In summary, $V_\eta(\overline{\mathbf{u}},\eta)$, is closely related to visibilities, $V_f(\overline{\mathbf{u}},f)$, the basic data blocks measured by an interferometer, as seen from equation~\ref{eqn:delay-transform}. It captures all the effects of EoR H{\sc i} signal corruption caused by foregrounds and the instrument. At the same time, it is very closely related to the sought power $|V(\overline{\mathbf{k}}_\perp,k_\parallel)|^2$ in $k$--modes containing critical information about spatial scales. For our analysis, we use an equivalent quantity, $V_\eta(\overline{\mathbf{x}},\eta) \equiv V_\eta(\lambda\overline{\mathbf{u}},\eta)$. For purposes of illustration, we often use a simplified quantity $|V_\eta(|\overline{\mathbf{x}}|,\eta)|$.

@@

\section{Instrument Parameters}\label{sec:instrument}

We use the 128-tile layout of the MWA \citep{bea12} and choose a center frequency of 185~MHz with a bandwidth of 30.72~MHz divided into 384 frequency channels each of width 80~kHz. The choice of center frequency translates to $z\approx 6.68$ for the 21~cm spin flip transition of H{\sc i}. The bandwidth was chosen to be close to the maximum instantaneous value of the MWA in order to maximize the resolution in delay coordinates and obtain detailed characteristics of delay spectra from foregrounds. We restrict our analysis to the innermost antenna spacings with $|\overline{\mathbf{x}}| \leq x_\textrm{max} = 200$~m, or $|\overline{\mathbf{u}}| \leq u_\textrm{max} = 123$ at a frequency of 185~MHz. These interferometers are responsive to transverse spatial structures on scales $\gtrsim 1/u_\textrm{max} \approx 28$\arcmin. Hereafter, we use the terms tile and antenna interchangeably to refer to the MWA tile.

The bandpass weights in the frequency window of the MWA are obtained through a 8-tap Polyphase Filter Bank (PFB) using a Kaiser window with the $\beta$ parameter set to 5. The bandpass weights consist of coarse frequency channels of width 1.28~MHz each of which contains 16 finer frequency channels of width 80~kHz. Our analysis, therefore, contains 24 coarse channels. As the data are subsequently bandpass calibrated, the bandpass weights are normalized to unity. But the edges of the coarse channels are noisy and aliased from adjacent coarse channels. Thus, we remove one fine channel from each side of the coarse channel edges from our data. This constitutes $W_f(f)$ in equation~\ref{eqn:obsvis}. The bandpass properties used in this paper are illustrated in figure~\ref{fig:bandpass}.

@@

Thermal noise in simulated visibilities is estimated assuming a system temperature of $T_\textrm{sys}=90$~K per polarization for all frequency channels for all interferometers throughout the course of observation. We take into account the increase in thermal uncertainty in frequency channels that results due to aforementioned bandpass correction.

\section{Observation Parameters}\label{sec:obsparms}

One of the primary targets for EoR observing with the MWA is a patch of sky centered around RA = 00h 00m 00.00s, Dec = -30\arcdeg$\,$00\arcmin$\,$00\farcs 0. This is one of the regions of low foreground emission within 27\arcdeg$\,$ from the zenith. The MWA tracks a patch of sky through antenna beams formed and steered electronically by controlling delay settings of an array of dipoles in a MWA tile. These beamformer delay settings can be changed only in discrete steps. Accordingly, observations were made in a ``drift and shift'' (D-S) mode. Figure~\ref{fig:pointings} shows the direction of pointing of MWA tiles during the course of the observation. The sky is allowed to drift for a certain period of time (usually $\sim 30$~mins) before the discrete shifts in the beamformer delay settings position the beam to be centered again on the same patch of sky. This process is repeated throughout the course of the observation $\approx 4.86$~hours. The LST of two snapshots we chose for our study are shown as vertical lines at -1.92~hours (wrapped by 360\arcdeg from 22.08~hours) and 0.09~hours which are hereafter denoted as {\it off--zenith} and {\it zenith} pointings respectively. These pointings (phased array power patterns) are centered at RA = 4\fdg 387, Dec = -29\fdg 86 and RA = -28\fdg 8, Dec = -26\fdg 701, whereas the visibilities themselves are phased to zenith corresponding to RA = -28\fdg 8, Dec = -26\fdg 701, and RA = 1\fdg 35, Dec = -26\fdg 701 respectively.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/custom_pointings.eps}
\caption{MWA tile beam pointing directions during the course of the observation. The x-axis refers to the Local Sidereal Time (LST) in hours. The axis on the left refers to longitudes, namely, Right Ascension (RA) and Hour Angle (HA) in degrees. Negative values of RA, HA, and LST are to be interpreted as having been wrapped around by 360\arcdeg or 24 hours. The axis on the right refers to the declination of the pointing direction. The RA, HA, and declination are plotted with solid, dashed and dotted line styles respectively. A ``drift and shift'' scheme is used. The sky is allowed to drift for $\sim 30$~mins before the beamformer delay settings change in a discrete step to center the beam around RA = 00h 00m 00.00s, Dec = -30\arcdeg$\,$00\arcmin$\,$00\farcs 0. Dot--dashed vertical lines show the LST of off-zenith and zenith pointings used in our study. \label{fig:pointings}}
\end{figure}

Throughout this paper, $\overline{\mathbf{u}}$ and $\overline{\mathbf{x}}$ are assumed to be on a coordinate system aligned with the local east, north (along local meridian) and upward directions at the MWA site. For the conventions of Fourier transform and its inverse we use in this paper, in accordance with equation~\ref{eqn:delay}, signals from the sky towards east and north are recorded with positive delays on eastward and northward oriented antenna spacings respectively. For instance, at the beginning of the observation, the Galactic center is in the westward sky just about to set. Signals from this direction are recorded with a negative delay on interferometers whose antenna spacings are oriented towards the east. Similarly, signals from directions eastward of the local meridian arrive with positive delays on eastward oriented antenna spacings.

For geometrical intuition, we restrict the orientation ($\theta_b$, measured anti-clockwise from East) of all antenna spacings to lie in the range $-67\fdg 5 \leq \theta_b < 112\fdg 5$. Interferometers with antenna spacings oriented in the other half-plane measure conjugate visibilities with delays of equal magnitude but of opposite sign and hence are ignored in our analysis.

\section{Data Analysis}\label{sec:data-analysis}

@@

We show results of delay spectrum from MWA data after deconvolution along delay axis in figure~\ref{fig:fhd_data}. Boundaries of the {\it foreground wedge} are also shown. The {\it off--zenith} pointing has emission notably higher than in the {\it zenith} pointing inside the wedge boundaries. This is shown later to be due to response of the interferometer array to the bright Galactic center and Galactic plane in the westward sky. Consequently, the contamination into the {\it EoR window} is also higher. Emission in {\it zenith} pointing is more centrally concentrated due to phasing of power pattern to the zenith, whereas the {\it off--zenith} pointing has a power pattern phased eastward of zenith and is also responsive to emission from angles far away from zenith.

@@

\item {\it Interferometer Orientation}: Since the spatial structure of our foreground model is not expected to be isotropic, we divide our interferometer antenna spacings by their orientation ($\theta_\textrm{b}$) measured counter-clockwise from East. We use the following bins: $-67\fdg 5\le \theta_\textrm{b} < -22\fdg 5$, $-22\fdg 5\le \theta_\textrm{b} < 22\fdg 5$, $22\fdg 5\le \theta_\textrm{b} < 67\fdg 5$, and $67\fdg 5\le \theta_\textrm{b} < 112\fdg 5$. The bin centers are oriented towards South-East, East, North-East, and North respectively. Since the delays registered at an interferometer depend on the orientation of its antenna spacing vector, binning by interferometer orientation allows us to attribute the delay spectrum to different sky directions.

@@

It may be noted that numerous features overlap at varying levels of significance as a result of various combinations of parameters. We assign the features to their predominant causes. Secondly, we have used noiseless cases to clearly illustrate the observed foreground signatures. With the addition of noise in the visibilities which is subject to observing time, some of the weaker features may not be as prominently visible. Since the foreground signatures are far too numerous and they reside in a vast volume of parameter space such as interferometer orientation, power pattern, patch of sky under observation, and instrumental configuration, we highlight only some examples of the most notable features observed in the delay spectrum of foreground emission.

@@

Figure~\ref{fig:pitchfork-baselines} shows delay spectra of three interferometers of different antenna spacings oriented northward during the {\it zenith} pointing. The diffuse, compact, and composite components are are shown as dotted, dashed, and solid lines respectively. The horizon delay limits are shown as vertical dot--dashed lines. The gray shaded area denotes the envelope of expected uncertainty in the delay spectrum. This uncertainty is dominated by the spread in spectral index of compact foreground objects inside the {\it foreground} wedge (between the horizon delay limits) proprotional to the delay spectrum from compact objects, and by thermal noise outside the horizon delay limits.

@@

We also note that increasing the antenna spacing results in progressively improving the resolution along delay axis by increasing the number of delay bins inside the {\it foreground wedge}. This improves the localization of foreground objects whose signatures are imprinted in the delay spectrum. For instance, there is an increase in the number of secondary peaks in the delay spectrum between zero--delay and horizon delay limits as antenna spacing increases from $\sim$84~m to $\sim$170~m. In this case, these correspond to sidelobes of the power pattern that lie between the main lobe and the horizon along the North--South direction. In interferometers with short spacings, due to relatively lower resolution along delay axis inside the {\it foreground wedge} and a consequent loss of localization of direction of foreground emission, these secondary peaks blend in with other major peaks and are not distinctly visible. These findings are consistent with dicussion in \S\ref{sec:csm-side-lobe}.

@@

\section{Screening of Severe Foreground Contaminated Interferometers}\label{sec:fg-grading}

@@

With a foreground model known {\it a priori} in which locations of very bright foreground objects such as the Galactic center or AGN are available, we can predict their response across antenna spacings as a function of observing parameters such as LST, power pattern, etc. Specifically, we can pick interferometers where these bright foreground objects will be visible at delays near horizon delay limits. One such example is already provided in our study. The Galactic center is the most dominant source of foreground contamination in the {\it off--zenith} pointing. Since it is located in the western sky, it is most distinctly seen near the negative horizon delay limit on eastward oriented antenna spacings in the delay spectrum (see figures~\ref{fig:fhd_data}, \ref{fig:sim_data} and \ref{fig:noiseless-dsm-delay-spectrum}, and \S\ref{sec:GC-east}). In fact these figures illustrate that deconvolution along delay axis is severly hampered by this extreme contamination, thereby leading to leakage beyond the {\it foreground wedge} into the {\it EoR window}. In practice, we could downweight these antenna spacings for the specific LST to mitigate the effects of spillover from such foreground contamination in EoR studies. With spatial scales of such objects known, besides downweighting based on antenna spacing orientation, we could also downweight based on antenna spacing distance. For instance, a bright compact object will excite interferometers uniformly across antenna spacings whereas an extended object will gradually fade with increase in antenna spacing.

This provides us with a very simple and yet effective tool in adding a layer of sophistication to mitigate effects of foreground contamination in EoR data analysis. % The success behind clear identification of severely foreground--contaminated interferometers based on delay spectrum technique depends on the localization of the source of foreground contamination and its strength.

@@

We also provide a simple and effective tool based on the delay spectrum technique to mitigate foreground contamination in EoR data analysis. It is possible to screen interferometers most affected by foreground contamination as a function of antenna spacing vector and time of observation, which can then be assigned optimal weights in the EoR data analysis chain.

>>>>

In summary, $V_\eta(\overline{\mathbf{u}},\eta)$, is closely related to visibilities, $V_f(\overline{\mathbf{u}},f)$, the basic data blocks measured by each baseline, as seen from equation~\ref{eqn:delay-transform}. It captures all the effects of EoR H{\sc i} signal corruption caused by foregrounds and the instrument. At the same time, it is very closely related to the sought power $|V(\overline{\mathbf{k}}_\perp,k_\parallel)|^2$ in $k$--modes containing critical information about spatial scales. For our analysis, we use an equivalent quantity, $V_\eta(\overline{\mathbf{x}},\eta) \equiv V_\eta(\lambda\overline{\mathbf{u}},\eta)$. For purposes of illustration, we often use a simplified quantity $|V_\eta(|\overline{\mathbf{x}}|,\eta)|$.

@@

\section{Murchison Widefield Array Observations}\label{sec:instrument}
\subsection{The MWA}
The MWA was completed in 2012, and after commissioning, began its Epoch of Reionization observing program  in 2013.  In its final configuration the MWA is a 128 tile interferometer capable of observing a 30MHz band anywhere between 80 to 300 MHz.  Each tile is a phased array of 16 bow-tie-shaped dipoles characterized by a grating response with a $\sim$ 25\arcdeg wide primary field of view and multiple secondary side lobes. The tile arrangement is a centrally condensed core of about 300m -there are many spacings in the range 5 to 50m- and a radial density that falls off as the inverse of the radius with the longest baselines at 3km \citep{bea12}. Here we focus on baselines shorter than 200m (spatial scales relevant to reionization) and a frequency range 170~MHz to 200~MHz.

One important feature of the MWA's measured spectrum is the shape of its bandpass. The band is divided coarsely into 24 chunks with each chunk weighted by a digital filter. The channels at the edges of these chunks have been flagged because they are known to possibly be contaminated by aliasing at a low level.  Any spectral flags, or other known weights constitutes $W_f(f)$ in equation~\ref{eqn:obsvis}. The bandpass properties used in this paper are illustrated in figure~\ref{fig:bandpass}.

@@

Thermal noise in simulated visibilities is estimated assuming a system temperature of $T_\textrm{sys}=90$~K per polarization for all frequency channels for all baselines throughout the course of observation. We take into account the increase in thermal uncertainty in frequency channels that results due to aforementioned bandpass correction. %XXX Since the bandpass correction is digital, shouldn't the noise stay flat after bandpass correction?

\subsection{Observation Parameters}\label{sec:obsparms}

One of the primary targets for EoR observing with the MWA is a patch of sky centered around RA = 00h 00m 00.00s, Dec = -30\arcdeg$\,$00\arcmin$\,$00\farcs 0. This is one of the regions of low foreground emission within 27\arcdeg$\,$ from the zenith. The MWA tracks a patch of sky through antenna beams formed and steered electronically by controlling delay settings of an array of dipoles in a MWA tile. These beamformer delay settings can be changed only in discrete steps.   The sky is allowed to drift for a certain period of time (usually $\sim 30$~mins) before the discrete shifts in the beamformer delay settings position the beam to be centered again on the same patch of sky. This process is repeated throughout the course of the observation $\approx 4.86$~hours. Figure~\ref{fig:pointings} illustrates how the apparent pointing oscillates around the desired pointing direction.

The LST of two snapshots we chose for our study are shown as vertical lines at -1.92~hours (i.e.  22.08~hours) and 0.09~hours which are hereafter denoted as {\it off--zenith} and {\it zenith} pointings respectively. These pointings (phased array power patterns) are centered at RA = 4\fdg 387, Dec = -29\fdg 86 and RA = -28\fdg 8, Dec = -26\fdg 701, whereas the visibilities themselves are phased to zenith corresponding to RA = -28\fdg 8, Dec = -26\fdg 701, and RA = 1\fdg 35, Dec = -26\fdg 701 respectively.

The observations reported here were observed, near the beginning of the campaign, on Aug 23, 2013.  This day has been chosen as a reference data set for comparison within the MWA EoR group. This day was selected on the basis of the low number of known system errors from among a long stretch of stable observing.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/custom_pointings.eps}
\caption{MWA tile beam pointing directions during the course of the observation. The x-axis refers to the Local Sidereal Time (LST) in hours. The axis on the left refers to longitudes, namely, Right Ascension (RA) and Hour Angle (HA) in degrees. Negative values of RA, HA, and LST are to be interpreted as having been wrapped around by 360\arcdeg or 24 hours. The axis on the right refers to the declination of the pointing direction. The RA, HA, and declination are plotted with solid, dashed and dotted line styles respectively. A ``drift and shift'' scheme is used. The sky is allowed to drift for $\sim 30$~mins before the beamformer delay settings change in a discrete step to center the beam around RA = 00h 00m 00.00s, Dec = -30\arcdeg$\,$00\arcmin$\,$00\farcs 0. Dot--dashed vertical lines show the LST of off-zenith and zenith pointings used in our study. \label{fig:pointings}}
\end{figure}
\subsection*{A note on baselines/delay spectra sign convention}
Throughout this paper, $\overline{\mathbf{u}}$ and $\overline{\mathbf{x}}$ are assumed to be on a coordinate system aligned with the local east, north (along local meridian) and upward directions at the MWA site. For the conventions of Fourier transform and its inverse we use in this paper, in accordance with equation~\ref{eqn:delay}, signals from the sky towards east and north are recorded with positive delays on eastward and northward oriented antenna spacings respectively. For instance, at the beginning of the observation, the Galactic center is in the westward sky just about to set. Signals from this direction are recorded with a negative delay on baselines whose antenna spacings are oriented towards the east. Similarly, signals from directions eastward of the local meridian arrive with positive delays on eastward oriented antenna spacings.

For geometrical intuition, we restrict the orientation ($\theta_b$, measured anti-clockwise from East) of all antenna spacings to lie in the range $-67\fdg 5 \leq \theta_b < 112\fdg 5$. baselines with antenna spacings oriented in the other half-plane measure conjugate visibilities with delays of equal magnitude but of opposite sign and hence are ignored in our analysis.

\subsection{Data Analysis}\label{sec:data-analysis}
The data are flagged for interference cite ao, removing 3\% of the data.  averaged in time and frequency to 2s, 40khz

@@

We show results of delay spectrum from MWA data after deconvolution along delay axis in figure~\ref{fig:fhd_data}. Boundaries of the {\it foreground wedge} are also shown. The {\it off--zenith} pointing has emission notably higher than in the {\it zenith} pointing inside the wedge boundaries. This is shown later to be due to response of the array to the bright Galactic center and Galactic plane in the westward sky. Consequently, the contamination into the {\it EoR window} is also higher. Emission in {\it zenith} pointing is more centrally concentrated due to phasing of power pattern to the zenith, whereas the {\it off--zenith} pointing has a power pattern phased eastward of zenith and is also responsive to emission from angles far away from zenith.

@@

\item {\it baseline Orientation}: Since the spatial structure of our foreground model is not expected to be isotropic, we divide our baselines by their orientation ($\theta_\textrm{b}$) measured counter-clockwise from East. We use the following bins: $-67\fdg 5\le \theta_\textrm{b} < -22\fdg 5$, $-22\fdg 5\le \theta_\textrm{b} < 22\fdg 5$, $22\fdg 5\le \theta_\textrm{b} < 67\fdg 5$, and $67\fdg 5\le \theta_\textrm{b} < 112\fdg 5$. The bin centers are oriented towards South-East, East, North-East, and North respectively. Because delays depend on baseline orientation, binning by baseline orientation allows us to match delay spectrum features to different sky directions.

@@

It may be noted that numerous features overlap at varying levels of significance as a result of various combinations of parameters. We assign the features to their predominant causes. Secondly, we have used noiseless cases to clearly illustrate the observed foreground signatures. With the addition of noise in the visibilities which is subject to observing time, some of the weaker features may not be as prominently visible. Since the foreground signatures are far too numerous and they reside in a vast volume of parameter space such as baseline orientation, power pattern, patch of sky under observation, and instrumental configuration, we highlight only some examples of the most notable features observed in the delay spectrum of foreground emission.

@@

Figure~\ref{fig:pitchfork-baselines} shows delay spectra of three baselines of different antenna spacings oriented northward during the {\it zenith} pointing. The diffuse, compact, and composite components are are shown as dotted, dashed, and solid lines respectively. The horizon delay limits are shown as vertical dot--dashed lines. The gray shaded area denotes the envelope of expected uncertainty in the delay spectrum. This uncertainty is dominated by the spread in spectral index of compact foreground objects inside the {\it foreground} wedge (between the horizon delay limits) proprotional to the delay spectrum from compact objects, and by thermal noise outside the horizon delay limits.

@@

We also note that increasing the antenna spacing results in progressively improving the resolution along delay axis by increasing the number of delay bins inside the {\it foreground wedge}. This improves the localization of foreground objects whose signatures are imprinted in the delay spectrum. For instance, there is an increase in the number of secondary peaks in the delay spectrum between zero--delay and horizon delay limits as antenna spacing increases from $\sim$84~m to $\sim$170~m. In this case, these correspond to sidelobes of the power pattern that lie between the main lobe and the horizon along the North--South direction. In baselines with short spacings, due to relatively lower resolution along delay axis inside the {\it foreground wedge} and a consequent loss of localization of direction of foreground emission, these secondary peaks blend in with other major peaks and are not distinctly visible. These findings are consistent with dicussion in \S\ref{sec:csm-side-lobe}.

@@

\section{Screening of Severe Foreground Contaminated baselines}\label{sec:fg-grading}

@@

With a foreground model known {\it a priori} in which locations of very bright foreground objects such as the Galactic center or AGN are available, we can predict their response across antenna spacings as a function of observing parameters such as LST, power pattern, etc. Specifically, we can pick baselines where these bright foreground objects will be visible at delays near horizon delay limits. One such example is already provided in our study. The Galactic center is the most dominant source of foreground contamination in the {\it off--zenith} pointing. Since it is located in the western sky, it is most distinctly seen near the negative horizon delay limit on eastward oriented antenna spacings in the delay spectrum (see figures~\ref{fig:fhd_data}, \ref{fig:sim_data} and \ref{fig:noiseless-dsm-delay-spectrum}, and \S\ref{sec:GC-east}). In fact these figures illustrate that deconvolution along delay axis is severly hampered by this extreme contamination, thereby leading to leakage beyond the {\it foreground wedge} into the {\it EoR window}. In practice, we could downweight these antenna spacings for the specific LST to mitigate the effects of spillover from such foreground contamination in EoR studies. With spatial scales of such objects known, besides downweighting based on antenna spacing orientation, we could also downweight based on antenna spacing distance. For instance, a bright compact object will excite baselines uniformly across antenna spacings whereas an extended object will gradually fade with increase in antenna spacing.

This provides us with a very simple and yet effective tool in adding a layer of sophistication to mitigate effects of foreground contamination in EoR data analysis. % The success behind clear identification of severely foreground--contaminated baselines based on delay spectrum technique depends on the localization of the source of foreground contamination and its strength.

@@

We also provide a simple and effective tool based on the delay spectrum technique to mitigate foreground contamination in EoR data analysis. It is possible to screen baselines most affected by foreground contamination as a function of antenna spacing vector and time of observation, which can then be assigned optimal weights in the EoR data analysis chain.
